---
title: "Getting Started with soilgraph"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with soilgraph}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 5,
  fig.height = 7,
  dpi = 150,
  out.width = "70%",
  fig.align = "center"
)
```

## Overview

**soilgraph** provides tools for structuring soil profile descriptions, exporting
them to JSON, and generating profile visualizations. A typical workflow is:

1. Prepare a structured table of horizon properties (or load a `.soil.json` file).
2. Build a `soil_profile` object.
3. Visualize the profile.
4. Export to the `.soil.json` format for archival and exchange.

```{r load-library}
library(soilgraph)
```

## Creating a structured horizon table

The recommended approach is to enter your horizon data as a data frame with
explicit columns. At minimum you need `Top` and `Bottom` depths; all other
fields are optional:

```{r structured-table}
horizons_df <- data.frame(
  Top = c(0, 18, 52),
  Bottom = c(18, 52, 95),
  Horizon = c("Ap", "Bt1", "Bt2"),
  Texture = c("silt loam", "clay loam", "clay"),
  Color = c("#5C4033", "#8A5A44", "#A66A4C"),
  Moisture = c("moist", "slightly moist", "dry"),
  BoundaryGrade = c("clear", "gradual", "diffuse"),
  BoundaryShape = c("smooth", "wavy", "irregular"),
  CoarseAbundance = c("few", "common", "many"),
  CoarseShape = c("subangular", "rounded", "subrounded"),
  CoarseGrade = c("weak", "moderate", "strong"),
  stringsAsFactors = FALSE
)
horizons_df
```

Both `PascalCase` (`BoundaryGrade`) and `snake_case` (`boundary_grade`) column
names are recognized.

## Building a soil profile object

`soil_profile_from_table()` converts a horizon table into a `soil_profile`
object, which is the primary data structure used by all plotting and export
functions:

```{r build-profile}
profile <- soil_profile_from_table(
  horizons_df,
  site_id = "pedon-001",
  classification = list(
    system = "Soil Taxonomy",
    taxon  = "Typic Hapludalf"
  ),
  metadata = list(country = "Chile", land_use = "cropland")
)
```

You can also build profiles manually from individual horizons:

```{r manual-profile}
h1 <- new_soil_horizon(0, 18,
  label = "Ap", color = "#5C4033",
  texture = "silt loam", boundary_grade = "clear", boundary_shape = "smooth"
)
h2 <- new_soil_horizon(18, 52,
  label = "Bt1", color = "#8A5A44",
  texture = "clay loam", boundary_grade = "gradual", boundary_shape = "wavy"
)
h3 <- new_soil_horizon(52, 95,
  label = "Bt2", color = "#A66A4C",
  texture = "clay", boundary_grade = "diffuse", boundary_shape = "irregular"
)

profile_manual <- new_soil_profile(
  site_id = "pedon-001",
  horizons = list(h1, h2, h3),
  classification = list(system = "Soil Taxonomy", taxon = "Typic Hapludalf")
)
```

Use `validate_soil_profile()` to check depth ordering and overlap constraints:

```{r validate}
validate_soil_profile(profile_manual)
```

## Plotting

### Basic profile

```{r plot-basic, fig.cap = "Basic soil profile visualization."}
plot_soil_profile(profile)
```

### Profile with coarse fragment markers

Point-based markers encode fragment type (shape), size (point size), and
cementation grade (alpha):

```{r plot-fragments, fig.cap = "Profile with coarse fragment point markers."}
plot_soil_profile_fragments(profile, seed = 42)
```

### Advanced visualization

The advanced renderer uses polygon-based fragments and irregular boundary lines:

```{r plot-advanced, fig.cap = "Advanced profile with polygon fragments, boundaries, and transition zones."}
plot_soil_profile_advanced(profile, seed = 42)
```

### Plotting directly from structured tables

For quick visualization you can go from table to plot in two steps:

```{r plot-from-table, fig.cap = "Plot from a structured table."}
quick_profile <- soil_profile_from_table(horizons_df, site_id = "pedon-001")
plot_soil_profile(quick_profile)
```

## Exporting and importing JSON

The `.soil.json` format stores profiles with all metadata in a standardized
schema:

```{r json-roundtrip}
# Write to a temporary file
tmp <- tempfile(fileext = ".soil.json")
write_soil_json(profile, tmp)

# Inspect the JSON structure
cat(readLines(tmp, n = 15), sep = "\n")
```

Read it back:

```{r json-read}
profile_reloaded <- read_soil_json(tmp)
profile_reloaded$site_id
profile_reloaded$classification$taxon
length(profile_reloaded$horizons)
```

## Reading bundled example data

The package ships with example files in `inst/extdata/`:

```{r bundled-data}
json_path <- system.file("extdata", "example.soil.json", package = "soilgraph")
example_profile <- read_soil_json(json_path)
plot_soil_profile(example_profile)
```

## Next steps

- Learn about the advanced graphical engines in `vignette("graphical-engines")`.
- Explore the soil data model and supported vocabulary in `vignette("soil-data-model")`.
- Browse a gallery of rendering examples in `vignette("rendering-examples")`.

## Migrating from the free-text parser

Prior versions of soilgraph offered `derive_soil_horizons()` and
`plot_soil_description()` to extract soil properties from free-text descriptions.
These functions are now **deprecated**. The recommended migration is
straightforward:

**Before (deprecated):**

```r
field_notes <- data.frame(
  Depth = c("0-18 cm", "18-52 cm"),
  description = c(
    "Ap dark brown silt loam moist clear smooth",
    "Bt1 reddish brown clay loam gradual wavy"
  )
)
profile <- soil_profile_from_table(field_notes, site_id = "pedon-001")
plot_soil_description(field_notes, site_id = "pedon-001")
```

**After (structured):**

```r
horizons_df <- data.frame(
  Top = c(0, 18),
  Bottom = c(18, 52),
  Horizon = c("Ap", "Bt1"),
  Texture = c("silt loam", "clay loam"),
  Color = c("#5C4033", "#8A5A44"),
  Moisture = c("moist", NA),
  BoundaryGrade = c("clear", "gradual"),
  BoundaryShape = c("smooth", "wavy"),
  stringsAsFactors = FALSE
)
profile <- soil_profile_from_table(horizons_df, site_id = "pedon-001")
plot_soil_profile(profile)
```

The structured approach is more explicit, more reliable, and lets you control
exactly which properties are recorded for each horizon.
