---
title: "Getting Started with soilgraph"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with soilgraph}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 5,
  fig.height = 7,
  dpi = 150,
  out.width = "70%",
  fig.align = "center"
)
```

## Overview

**soilgraph** provides tools for structuring soil profile descriptions, exporting
them to JSON, and generating profile visualizations. A typical workflow is:

1. Prepare a two-column field description table (**Depth** + **description**).
2. Parse descriptions into a structured soil profile.
3. Visualize the profile.
4. Export to the `.soil.json` format for archival and exchange.

```{r load-library}
library(soilgraph)
```

## Creating a field description table

Field notes are represented as a data frame with a `Depth` column containing depth
ranges (e.g. `"0-18 cm"`) and a `description` column with natural language
descriptions:

```{r field-notes}
field_notes <- data.frame(
  Depth = c("0-18 cm", "18-52 cm", "52-95 cm"),
  description = c(
    "Ap dark brown silt loam moist clear smooth few subangular weak fragments with granular structure",
    "Bt1 reddish brown clay loam slightly moist gradual wavy common rounded moderate fragments with clay films",
    "Bt2 brown clay dry diffuse irregular many subrounded strong fragments with strong blocky structure"
  )
)
field_notes
```

## Parsing descriptions

`derive_soil_horizons()` extracts horizon properties (texture, color, moisture,
boundary, coarse fragments) from the description text and returns a tidy table:

```{r derive}
horizons <- derive_soil_horizons(field_notes)
horizons
```

The parser recognizes two input modes:

- **Depth ranges**: `"0-18 cm"`, `"18-52 cm"`, `"52-95 cm"`
- **Top depths only**: `"0"`, `"18"`, `"52"` — requires a `profile_bottom` argument.

```{r top-depth-mode}
top_depth_notes <- data.frame(
  Depth = c("0", "18", "52"),
  description = c(
    "Ap dark brown silt loam moist clear smooth",
    "Bt1 reddish brown clay loam gradual wavy",
    "Bt2 brown clay dry diffuse irregular"
  )
)

horizons_top <- derive_soil_horizons(top_depth_notes, profile_bottom = 95)
horizons_top[, c("Horizon", "Top", "Bottom", "Texture", "Color")]
```

## Building a soil profile object

`soil_profile_from_table()` converts a field notes table into a `soil_profile`
object, which is the primary data structure used by all plotting and export
functions:

```{r build-profile}
profile <- soil_profile_from_table(
  field_notes,
  site_id = "pedon-001",
  classification = list(
    system = "Soil Taxonomy",
    taxon  = "Typic Hapludalf"
  ),
  metadata = list(country = "Chile", land_use = "cropland")
)
```

You can also build profiles manually from individual horizons:

```{r manual-profile}
h1 <- new_soil_horizon(0, 18, label = "Ap", color = "#5C4033",
  texture = "silt loam", boundary_grade = "clear", boundary_shape = "smooth")
h2 <- new_soil_horizon(18, 52, label = "Bt1", color = "#8A5A44",
  texture = "clay loam", boundary_grade = "gradual", boundary_shape = "wavy")
h3 <- new_soil_horizon(52, 95, label = "Bt2", color = "#A66A4C",
  texture = "clay", boundary_grade = "diffuse", boundary_shape = "irregular")

profile_manual <- new_soil_profile(
  site_id = "pedon-001",
  horizons = list(h1, h2, h3),
  classification = list(system = "Soil Taxonomy", taxon = "Typic Hapludalf")
)
```

Use `validate_soil_profile()` to check depth ordering and overlap constraints:

```{r validate}
validate_soil_profile(profile_manual)
```

## Plotting

### Basic profile

```{r plot-basic, fig.cap = "Basic soil profile visualization."}
plot_soil_profile(profile)
```

### Profile with coarse fragment markers

Point-based markers encode fragment type (shape), size (point size), and
cementation grade (alpha):

```{r plot-fragments, fig.cap = "Profile with coarse fragment point markers."}
plot_soil_profile_fragments(profile, seed = 42)
```

### Advanced visualization

The advanced renderer uses polygon-based fragments and irregular boundary lines:

```{r plot-advanced, fig.cap = "Advanced profile with polygon fragments, boundaries, and transition zones."}
plot_soil_profile_advanced(profile, seed = 42)
```

### Plotting directly from field notes

For quick visualization without creating objects manually:

```{r plot-description, fig.cap = "Plot directly from field description table."}
plot_soil_description(field_notes, site_id = "pedon-001")
```

## Exporting and importing JSON

The `.soil.json` format stores profiles with all metadata in a standardized
schema:

```{r json-roundtrip}
# Write to a temporary file
tmp <- tempfile(fileext = ".soil.json")
write_soil_json(profile, tmp)

# Inspect the JSON structure
cat(readLines(tmp, n = 15), sep = "\n")
```

Read it back:

```{r json-read}
profile_reloaded <- read_soil_json(tmp)
profile_reloaded$site_id
profile_reloaded$classification$taxon
length(profile_reloaded$horizons)
```

## Reading bundled example data

The package ships with example files in `inst/extdata/`:

```{r bundled-data}
json_path <- system.file("extdata", "example.soil.json", package = "soilgraph")
example_profile <- read_soil_json(json_path)
plot_soil_profile(example_profile)
```

## Next steps

- Learn about the advanced graphical engines in `vignette("graphical-engines")`.
- Explore the soil data model and supported vocabulary in `vignette("soil-data-model")`.
- Browse a gallery of rendering examples in `vignette("rendering-examples")`.
