---
title: "Graphical Engines: Advanced Visualization"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Graphical Engines: Advanced Visualization}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 5,
  fig.height = 7,
  dpi = 150,
  out.width = "70%",
  fig.align = "center"
)
```

## Overview

soilgraph ships with three **graphical engines** that convert soil morphological
properties into visual attributes:

| Engine | Encodes | Visual output |
|---|---|---|
| **Coarse Fragment Engine** | type, size, abundance, cementation grade | Irregular polygons |
| **Horizon Boundary Engine** | shape (topography), grade (distinctness) | Irregular boundary lines |
| **Visualization Utilities** | color, palette, theme, encoding | Themes and helpers |


The primary entry point is `plot_soil_profile_advanced()`, which combines all
three engines in a single call.

```{r load}
library(soilgraph)
```

## Setting up a reference profile

We will reuse this 3-horizon profile throughout the article:

```{r reference-profile}
profile <- new_soil_profile(
  site_id = "Engine-Demo",
  classification = list(system = "Soil Taxonomy", taxon = "Typic Hapludalf"),
  horizons = list(
    new_soil_horizon(
      top = 0, bottom = 18, label = "Ap",
      texture = "silt loam", color = "#5C4033",
      boundary_grade = "clear", boundary_shape = "smooth",
      coarse_type = "gravel", coarse_abundance = "few",
      coarse_grade = "weak", coarse_size = "fine"
    ),
    new_soil_horizon(
      top = 18, bottom = 52, label = "Bt1",
      texture = "clay loam", color = "#8A5A44",
      boundary_grade = "gradual", boundary_shape = "wavy",
      coarse_type = "cobble", coarse_abundance = "common",
      coarse_grade = "moderate", coarse_size = "medium"
    ),
    new_soil_horizon(
      top = 52, bottom = 95, label = "Bt2",
      texture = "clay", color = "#A66A4C",
      boundary_grade = "diffuse", boundary_shape = "irregular",
      coarse_type = "stone", coarse_abundance = "many",
      coarse_grade = "strong", coarse_size = "coarse"
    )
  )
)
```

## Basic vs. advanced rendering

Compare the three rendering modes:

```{r basic-plot, fig.cap = "Basic profile — colored rectangles with labels."}
plot_soil_profile(profile)
```

```{r point-fragments, fig.cap = "Point-based fragment markers."}
plot_soil_profile_fragments(profile, seed = 42)
```

```{r advanced-full, fig.cap = "Advanced rendering with polygon fragments, boundary lines, and transition zones."}
plot_soil_profile_advanced(profile, seed = 42)
```

## Toggling engine layers

You can enable or disable each engine independently:

```{r fragments-only, fig.cap = "Polygon fragments only — boundaries disabled."}
plot_soil_profile_advanced(
  profile,
  show_fragments = TRUE,
  show_boundaries = FALSE,
  show_transition_zones = FALSE,
  seed = 42
)
```

```{r boundaries-only, fig.cap = "Boundaries and transition zones only — fragments disabled."}
plot_soil_profile_advanced(
  profile,
  show_fragments = FALSE,
  show_boundaries = TRUE,
  show_transition_zones = TRUE,
  seed = 42
)
```

## Coarse Fragment Engine

### Fragment types

Six fragment types are supported, each with a distinct polygon shape:

| Type | Vertices | Aspect ratio | Irregularity |
|------|----------|-------------|-------------|
| gravel | 6 | 1.0 | 0.20 |
| cobble | 7 | 1.1 | 0.18 |
| stone | 5 | 1.2 | 0.25 |
| boulder | 8 | 1.0 | 0.15 |
| channer | 4 | 2.0 | 0.10 |
| flagstone | 4 | 2.5 | 0.08 |

```{r fragment-types, fig.height = 9, fig.cap = "All fragment types at equal abundance."}
fragment_types <- c("gravel", "cobble", "stone", "boulder", "channer", "flagstone")

type_profile <- new_soil_profile(
  site_id = "Fragment-Types",
  horizons = lapply(seq_along(fragment_types), function(i) {
    new_soil_horizon(
      top = (i - 1) * 15, bottom = i * 15,
      label = toupper(fragment_types[i]),
      color = "#8B7355",
      coarse_type = fragment_types[i],
      coarse_abundance = "common",
      coarse_grade = "moderate",
      coarse_size = "medium",
      boundary_shape = "smooth",
      boundary_grade = "clear"
    )
  })
)

plot_soil_profile_advanced(type_profile, show_boundaries = FALSE, seed = 42)
```

### Fragment abundance

The abundance property controls the **number** of fragments rendered in each
horizon. Five levels are supported:

| Abundance | Approximate count |
|-----------|-------------------|
| very few | 6 |
| few | 10 |
| common | 18 |
| many | 30 |
| abundant | 42 |

```{r fragment-abundance, fig.height = 8, fig.cap = "Fragment abundance gradient."}
abundances <- c("very few", "few", "common", "many", "abundant")

abundance_profile <- new_soil_profile(
  site_id = "Abundance-Gradient",
  horizons = lapply(seq_along(abundances), function(i) {
    new_soil_horizon(
      top = (i - 1) * 15, bottom = i * 15,
      label = abundances[i],
      color = "#8B7355",
      coarse_type = "gravel",
      coarse_abundance = abundances[i],
      coarse_grade = "moderate",
      coarse_size = "medium",
      boundary_shape = "smooth",
      boundary_grade = "clear"
    )
  })
)

plot_soil_profile_advanced(abundance_profile, show_boundaries = FALSE, seed = 42)
```

### Fragment cementation grade

The grade property is mapped to **alpha** (transparency), representing
cementation from very weak (translucent) to very strong (opaque):

| Grade | Alpha |
|-------|-------|
| very weak | 0.30 |
| weak | 0.45 |
| moderate | 0.65 |
| strong | 0.82 |
| very strong | 0.95 |

```{r fragment-grade, fig.height = 8, fig.cap = "Fragment cementation grade gradient."}
grades <- c("very weak", "weak", "moderate", "strong", "very strong")

grade_profile <- new_soil_profile(
  site_id = "Grade-Showcase",
  horizons = lapply(seq_along(grades), function(i) {
    new_soil_horizon(
      top = (i - 1) * 15, bottom = i * 15,
      label = grades[i],
      color = "#8B7355",
      coarse_type = "gravel",
      coarse_abundance = "common",
      coarse_grade = grades[i],
      coarse_size = "medium",
      boundary_shape = "smooth",
      boundary_grade = "clear"
    )
  })
)

plot_soil_profile_advanced(grade_profile, show_boundaries = FALSE, seed = 42)
```

### Fragment size classes

Size is mapped to **polygon scale**:

```{r fragment-size, fig.height = 9, fig.cap = "Fragment size class comparison."}
sizes <- c("very fine", "fine", "small", "medium", "coarse", "large", "very coarse")

size_profile <- new_soil_profile(
  site_id = "Size-Showcase",
  horizons = lapply(seq_along(sizes), function(i) {
    new_soil_horizon(
      top = (i - 1) * 12, bottom = i * 12,
      label = sizes[i],
      color = "#8B7355",
      coarse_type = "gravel",
      coarse_abundance = "common",
      coarse_grade = "moderate",
      coarse_size = sizes[i],
      boundary_shape = "smooth",
      boundary_grade = "clear"
    )
  })
)

plot_soil_profile_advanced(size_profile, show_boundaries = FALSE, seed = 42)
```

## Horizon Boundary Engine

### Boundary shapes (topography)

Shape determines the **distortion pattern** of the boundary line:

| Shape | Pattern | Distortion |
|-------|---------|------------|
| smooth | flat | 0.0 |
| wavy | sine-like | 0.6 |
| irregular | multi-frequency | 0.8 |
| broken | segment offsets | 1.0 |
| discontinuous | gaps | 0.9 |

### Boundary grades (distinctness)

Grade controls **line thickness** and **dash pattern**:

| Grade | Line width | Line type |
|-------|-----------|-----------|
| abrupt | 0.8 | solid |
| clear | 0.5 | solid |
| gradual | 0.35 | dashed |
| diffuse | 0.2 | dotted |

```{r boundary-properties, fig.height = 8, fig.cap = "Boundary grade and shape combinations."}
boundary_profile <- new_soil_profile(
  site_id = "Boundary-Types",
  horizons = list(
    new_soil_horizon(0, 15,
      label = "Abrupt / Smooth", color = "#A0826D",
      boundary_grade = "abrupt", boundary_shape = "smooth"
    ),
    new_soil_horizon(15, 30,
      label = "Clear / Wavy", color = "#9B7D66",
      boundary_grade = "clear", boundary_shape = "wavy"
    ),
    new_soil_horizon(30, 50,
      label = "Gradual / Irregular", color = "#96795F",
      boundary_grade = "gradual", boundary_shape = "irregular"
    ),
    new_soil_horizon(50, 75,
      label = "Diffuse / Broken", color = "#917558",
      boundary_grade = "diffuse", boundary_shape = "broken"
    )
  )
)

plot_soil_profile_advanced(
  boundary_profile,
  show_fragments = FALSE,
  show_boundaries = TRUE,
  show_transition_zones = TRUE,
  seed = 42
)
```

### Transition zones

Gradual and diffuse boundaries generate transparent transition zones that blend
adjacent horizon colors. Enable them with `show_transition_zones = TRUE`
(the default).

## Seed reproducibility

All random elements (fragment placement, polygon perturbation, boundary noise)
are seeded. The same `seed` value always produces the same plot:

```{r seed, fig.width = 5, fig.height = 7, fig.cap = "Deterministic rendering with seed = 42."}
p1 <- plot_soil_profile_advanced(profile, seed = 42)
p2 <- plot_soil_profile_advanced(profile, seed = 42)

# Both plots are identical
identical(
  ggplot2::ggplot_build(p1)$data,
  ggplot2::ggplot_build(p2)$data
)
```

## Low-level engine functions

For custom visualizations you can call the engines directly. These functions
are internal (not exported), so use the `:::` operator:

```{r low-level}
# Prepare horizon data
horizon_data <- soilgraph:::build_horizon_plot_data(profile)
head(horizon_data[, c("label", "top", "bottom", "fill", "boundary_shape", "coarse_type")])

# Generate boundary paths
boundary_paths <- soilgraph:::build_boundary_paths(horizon_data, seed = 42)
head(boundary_paths)

# Generate fragment polygons
fragment_polys <- soilgraph:::build_fragment_polygons(profile, seed = 42)
head(fragment_polys)
```

These data frames can be used with standard ggplot2 geoms for fully
customized visualizations.

## Summary

The three graphical engines translate soil morphological properties into
consistent visual encodings:

- **Fragments**: type → shape, size → scale, abundance → count, grade → alpha
- **Boundaries**: shape → distortion pattern, grade → line weight & dash
- **Utilities**: color validation, palette generation, contrast detection, theming

All engines are deterministic (seeded), composable (toggle independently), and
built on ggplot2 layers.
