---
title: "Rendering Examples Gallery"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Rendering Examples Gallery}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 5,
  fig.height = 7,
  dpi = 150,
  out.width = "70%",
  fig.align = "center"
)
```

This gallery shows a variety of soil profile renderings using the soilgraph
graphical engines. Each example includes the code and rendered plot.

```{r load}
library(soilgraph)
```

## 1. Typical Hapludalf (Chile)

A 3-horizon profile with increasing fragment content and progressively
less-distinct boundaries with depth:

```{r hapludalf, fig.cap = "Typic Hapludalf — silt loam over clay with increasing coarse fragments."}
profile_1 <- new_soil_profile(
  site_id = "Hapludalf-CL",
  classification = list(system = "Soil Taxonomy", taxon = "Typic Hapludalf"),
  horizons = list(
    new_soil_horizon(0, 18,
      label = "Ap", color = "#5C4033",
      texture = "silt loam", boundary_grade = "clear", boundary_shape = "smooth",
      coarse_type = "gravel", coarse_abundance = "few",
      coarse_grade = "weak", coarse_size = "fine"
    ),
    new_soil_horizon(18, 52,
      label = "Bt1", color = "#8A5A44",
      texture = "clay loam", boundary_grade = "gradual", boundary_shape = "wavy",
      coarse_type = "cobble", coarse_abundance = "common",
      coarse_grade = "moderate", coarse_size = "medium"
    ),
    new_soil_horizon(52, 95,
      label = "Bt2", color = "#A66A4C",
      texture = "clay", boundary_grade = "diffuse", boundary_shape = "irregular",
      coarse_type = "stone", coarse_abundance = "many",
      coarse_grade = "strong", coarse_size = "coarse"
    )
  )
)

plot_soil_profile_advanced(profile_1, seed = 42)
```

## 2. Deep sandy profile

A deep sandy soil with few fragments and smooth polygon edges — a clean,
minimal rendering:

```{r sandy, fig.height = 8, fig.cap = "Deep sandy profile with smooth boundaries."}
profile_2 <- new_soil_profile(
  site_id = "Sandy-Profile",
  classification = list(system = "Soil Taxonomy", taxon = "Typic Quartzipsamment"),
  horizons = list(
    new_soil_horizon(0, 20,
      label = "A", color = "#C2A878",
      texture = "sand", boundary_grade = "clear", boundary_shape = "smooth"
    ),
    new_soil_horizon(20, 55,
      label = "C1", color = "#D4BC8E",
      texture = "sand", boundary_grade = "gradual", boundary_shape = "smooth"
    ),
    new_soil_horizon(55, 100,
      label = "C2", color = "#E0CCA0",
      texture = "sand", boundary_grade = "diffuse", boundary_shape = "smooth"
    ),
    new_soil_horizon(100, 150,
      label = "C3", color = "#E8D8B0",
      texture = "loamy sand", boundary_grade = "diffuse", boundary_shape = "smooth"
    )
  )
)

plot_soil_profile_advanced(
  profile_2,
  show_fragments = FALSE,
  show_boundaries = TRUE,
  show_transition_zones = TRUE,
  seed = 42
)
```

## 3. Stony colluvial soil

Abundant boulders and channers in a thin dark soil over rock debris:

```{r colluvial, fig.height = 6, fig.cap = "Stony colluvial soil with abundant coarse fragments."}
profile_3 <- new_soil_profile(
  site_id = "Colluvial-01",
  classification = list(system = "Field description", taxon = NULL),
  horizons = list(
    new_soil_horizon(0, 12,
      label = "A", color = "#3D2B1F",
      texture = "loam", boundary_grade = "abrupt", boundary_shape = "irregular",
      coarse_type = "channer", coarse_abundance = "common",
      coarse_grade = "strong", coarse_size = "small"
    ),
    new_soil_horizon(12, 45,
      label = "C", color = "#6B5A4E",
      texture = "sandy loam", boundary_grade = "clear", boundary_shape = "broken",
      coarse_type = "boulder", coarse_abundance = "abundant",
      coarse_grade = "very strong", coarse_size = "very coarse"
    )
  )
)

plot_soil_profile_advanced(profile_3, seed = 42)
```

## 4. Fragment type comparison

All six fragment types at equal abundance and grade, side by side:

```{r fragment-types, fig.height = 9, fig.cap = "All six coarse fragment types."}
frag_types <- c("gravel", "cobble", "stone", "boulder", "channer", "flagstone")

profile_4 <- new_soil_profile(
  site_id = "Fragment-Types",
  horizons = lapply(seq_along(frag_types), function(i) {
    new_soil_horizon(
      top = (i - 1) * 15, bottom = i * 15,
      label = frag_types[i],
      color = "#8B7355",
      coarse_type = frag_types[i],
      coarse_abundance = "common",
      coarse_grade = "moderate",
      coarse_size = "medium",
      boundary_shape = "smooth",
      boundary_grade = "clear"
    )
  })
)

plot_soil_profile_advanced(profile_4, show_boundaries = FALSE, seed = 42)
```

## 5. Boundary showcase

Four grade x shape combinations showing edge distortion and transition ribbons:

```{r boundary-showcase, fig.height = 8, fig.cap = "Boundary grade and shape demonstration."}
profile_5 <- new_soil_profile(
  site_id = "Boundary-Demo",
  horizons = list(
    new_soil_horizon(0, 15,
      label = "Abrupt\nSmooth", color = "#A0826D",
      boundary_grade = "abrupt", boundary_shape = "smooth"
    ),
    new_soil_horizon(15, 30,
      label = "Clear\nWavy", color = "#9B7D66",
      boundary_grade = "clear", boundary_shape = "wavy"
    ),
    new_soil_horizon(30, 50,
      label = "Gradual\nIrregular", color = "#96795F",
      boundary_grade = "gradual", boundary_shape = "irregular"
    ),
    new_soil_horizon(50, 75,
      label = "Diffuse\nBroken", color = "#917558",
      boundary_grade = "diffuse", boundary_shape = "broken"
    )
  )
)

plot_soil_profile_advanced(
  profile_5,
  show_fragments = FALSE,
  show_boundaries = TRUE,
  show_transition_zones = TRUE,
  seed = 42
)
```

## 6. Flat-edge fallback

Disable shaped contacts to render flat horizon edges in advanced mode:

```{r flat-fallback, fig.cap = "Advanced renderer with shaped boundaries disabled (flat fallback)."}
plot_soil_profile_advanced(
  profile_5,
  show_fragments = FALSE,
  show_boundaries = FALSE,
  show_transition_zones = TRUE,
  seed = 42
)
```

## 7. Fragment abundance gradient

From very few to abundant fragments in identical horizons:

```{r abundance-gradient, fig.height = 8, fig.cap = "Fragment abundance increasing with depth."}
abundances <- c("very few", "few", "common", "many", "abundant")

profile_6 <- new_soil_profile(
  site_id = "Abundance-Gradient",
  horizons = lapply(seq_along(abundances), function(i) {
    new_soil_horizon(
      top = (i - 1) * 15, bottom = i * 15,
      label = abundances[i],
      color = "#8B7355",
      coarse_type = "gravel",
      coarse_abundance = abundances[i],
      coarse_grade = "moderate",
      coarse_size = "medium",
      boundary_shape = "smooth",
      boundary_grade = "clear"
    )
  })
)

plot_soil_profile_advanced(profile_6, show_boundaries = FALSE, seed = 42)
```

## 8. From field notes to advanced plot

End-to-end workflow — field notes parsed and rendered with advanced engines:

```{r field-to-plot, fig.cap = "Field description parsed and rendered with all engines."}
field_notes <- data.frame(
  Depth = c("0-18 cm", "18-52 cm", "52-95 cm"),
  description = c(
    "Ap dark brown silt loam moist clear smooth few subangular weak fragments",
    "Bt1 reddish brown clay loam slightly moist gradual wavy common rounded moderate fragments",
    "Bt2 brown clay dry diffuse irregular many subrounded strong fragments"
  )
)

profile_7 <- soil_profile_from_table(
  field_notes,
  site_id = "Field-to-Plot",
  classification = list(system = "Soil Taxonomy", taxon = "Typic Hapludalf")
)

plot_soil_profile_advanced(profile_7, seed = 42)
```

## 9. Seed variation

Same profile rendered with three different seeds — structures differ but properties
are preserved:

```{r seed-variation, fig.width = 5, fig.height = 7, fig.cap = "Seed = 1 — first random layout."}
plot_soil_profile_advanced(profile_1, seed = 1)
```

```{r seed-42, fig.width = 5, fig.height = 7, fig.cap = "Seed = 42 — different fragment placement."}
plot_soil_profile_advanced(profile_1, seed = 42)
```

```{r seed-99, fig.width = 5, fig.height = 7, fig.cap = "Seed = 99 — yet another layout."}
plot_soil_profile_advanced(profile_1, seed = 99)
```
