---
title: "Soil Description Data Model"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Soil Description Data Model}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Overview

soilgraph uses a two-level data model: **profiles** contain **horizons**. Each
level carries its own set of properties. This article documents the supported
fields, vocabularies, and the `.soil.json` storage format.

```{r load}
if (!requireNamespace("soilgraph", quietly = TRUE)) {
  if (!requireNamespace("devtools", quietly = TRUE)) {
    stop(
      "Package 'soilgraph' is not installed. Install it, or install 'devtools' ",
      "and render from the repository so the package can be loaded from source."
    )
  }

  pkg_root_candidates <- c(".", "..")
  pkg_root <- pkg_root_candidates[file.exists(file.path(pkg_root_candidates, "DESCRIPTION"))][1]

  if (is.na(pkg_root)) {
    stop("Could not find package root containing DESCRIPTION.")
  }

  devtools::load_all(path = pkg_root, quiet = TRUE)
}

library(soilgraph)
```

## Profile-level fields

A `soil_profile` object stores site-level metadata alongside a list of horizons:
nbsp;
| Field | Type | Required | Description |
|-------|------|----------|-------------|
| `schema_version` | string | auto | Schema version (currently `"0.1.0"`) |
| `site_id` | string | yes | Unique site identifier |
| `classification$system` | string | yes | Classification system (`"Soil Taxonomy"`, `"WRB"`) |
| `classification$taxon` | string | no | Taxon name (e.g. `"Typic Hapludalf"`) |
| `metadata` | named list | no | Arbitrary key-value pairs (country, land use, slope, etc.) |

```{r profile-fields}
profile <- new_soil_profile(
  site_id = "pedon-042",
  horizons = list(
    new_soil_horizon(0, 25, label = "A"),
    new_soil_horizon(25, 60, label = "Bw")
  ),
  classification = list(system = "WRB", taxon = "Haplic Cambisol"),
  metadata = list(country = "Chile", slope_percent = 8, land_use = "forest")
)

profile$site_id
profile$classification
profile$metadata
```

## Horizon-level fields

Each `soil_horizon` carries morphological properties. The table below lists all
supported fields:

### Core morphology

| Field | Type | Description |
|-------|------|-------------|
| `top` | numeric | Top depth (cm) — **required** |
| `bottom` | numeric | Bottom depth (cm) — **required** |
| `label` | string | Horizon designation (e.g. `"Ap"`, `"Bt1"`) |
| `texture` | string | Texture class (`"silt loam"`, `"clay"`, etc.) |
| `color` | string | Display color — hex code or R colour name |
| `moisture` | string | Moisture state (`"dry"`, `"moist"`, `"wet"`) |
| `notes` | string | Free-text notes |

### Horizon boundary

| Field | Type | Recognized values |
|-------|------|-------------------|
| `boundary_grade` | string | `abrupt`, `clear`, `gradual`, `diffuse` |
| `boundary_shape` | string | `smooth`, `wavy`, `irregular`, `broken`, `discontinuous` |

### Coarse fragments

| Field | Type | Recognized values |
|-------|------|-------------------|
| `coarse_abundance` | string | `very few`, `few`, `common`, `many`, `abundant` |
| `coarse_shape` | string | `angular`, `subangular`, `subrounded`, `rounded`, `flat` |
| `coarse_grade` | string | `very weak`, `weak`, `moderate`, `strong`, `very strong` |
| `coarse_type` | string | `gravel`, `cobble`, `stone`, `boulder`, `channer`, `flagstone` |
| `coarse_size` | string | `very fine`, `fine`, `small`, `medium`, `coarse`, `large`, `very coarse` |
| `coarse_color` | string | Fragment color — hex code or R colour name |
| `coarse_percent` | numeric | Volume percentage (0–100) |

```{r horizon-example}
h <- new_soil_horizon(
  top = 0, bottom = 18,
  label = "Ap",
  texture = "silt loam",
  color = "#5C4033",
  moisture = "moist",
  boundary_grade = "clear",
  boundary_shape = "smooth",
  coarse_abundance = "few",
  coarse_shape = "subangular",
  coarse_grade = "weak",
  coarse_type = "gravel",
  coarse_size = "fine",
  coarse_color = "#1B1B1B",
  coarse_percent = 5,
  notes = "Granular structure, abundant roots"
)
str(h, max.level = 1)
```

## Parsing natural language descriptions (deprecated)

> **Note:** `derive_soil_horizons()` and the free-text parser are deprecated.
> The recommended approach is to provide structured data directly via
> `soil_profile_from_table()` with explicit columns. See
> `vignette("getting-started")` for the migration path.

`derive_soil_horizons()` extracted structured data from free-text field
descriptions. The parser used keyword dictionaries to recognize properties.
This section is retained for reference only.

### Supported input formats

**Depth ranges** (e.g. `"0-18 cm"`):

```{r parse-range}
notes_range <- data.frame(
  Depth = c("0-18 cm", "18-52 cm", "52-95 cm"),
  description = c(
    "Ap dark brown silt loam moist clear smooth few subangular weak fragments",
    "Bt1 reddish brown clay loam slightly moist gradual wavy common rounded moderate fragments",
    "Bt2 brown clay dry diffuse irregular many subrounded strong fragments"
  )
)

result <- derive_soil_horizons(notes_range)
result[, c("Horizon", "Top", "Bottom", "Texture", "Color", "BoundaryGrade")]
```

**Top depths only** (requires `profile_bottom`):

```{r parse-top}
notes_top <- data.frame(
  Depth = c("0", "18", "52"),
  description = c(
    "Ap dark brown silt loam moist",
    "Bt1 reddish brown clay loam slightly moist",
    "Bt2 brown clay dry"
  )
)

result_top <- derive_soil_horizons(notes_top, profile_bottom = 95)
result_top[, c("Horizon", "Top", "Bottom")]
```

### Recognized vocabulary

The parser matches against built-in dictionaries. Here are the main categories:

**Texture classes**: sandy clay, sandy clay loam, sandy loam, loamy sand, silt
loam, silty clay, silty clay loam, clay loam, clay, loam, silt, sand.

**Moisture states**: dry, slightly moist, moist, wet.

**Color names**: dark brown, reddish brown, brown, yellowish brown, olive brown,
gray, dark gray, dark reddish brown, red, yellowish red, dark yellowish brown,
pale brown, light brownish gray, very dark gray, very dark brown, black, white.

**Boundary grades**: abrupt, clear, gradual, diffuse.

**Boundary shapes**: smooth, wavy, irregular, broken, discontinuous.

**Coarse fragment terms**: abundance (very few, few, common, many, abundant),
shape (angular, subangular, subrounded, rounded, flat), grade (very weak, weak,
moderate, strong, very strong).

## The `.soil.json` format

Profiles are serialized to a standardized JSON schema. The format includes the
full profile structure with all horizons and metadata.

```{r json-format}
profile_full <- soil_profile_from_table(
  notes_range,
  site_id = "pedon-001",
  classification = list(system = "Soil Taxonomy", taxon = "Typic Hapludalf"),
  metadata = list(country = "Chile")
)

tmp <- tempfile(fileext = ".soil.json")
write_soil_json(profile_full, tmp)
cat(readLines(tmp), sep = "\n")
```

### Example `.soil.json` structure

The bundled example file demonstrates the complete schema:

```{r bundled-json}
json_path <- system.file("extdata", "example.soil.json", package = "soilgraph")
example <- read_soil_json(json_path)

cat("Site ID:", example$site_id, "\n")
cat("System:", example$classification$system, "\n")
cat("Taxon:", example$classification$taxon, "\n")
cat("Horizons:", length(example$horizons), "\n")
cat("Total depth:", example$horizons[[length(example$horizons)]]$bottom, "cm\n")
```

### Round-trip fidelity

Writing and reading back preserves all fields:

```{r roundtrip}
tmp2 <- tempfile(fileext = ".soil.json")
write_soil_json(example, tmp2)
reloaded <- read_soil_json(tmp2)

identical(example$site_id, reloaded$site_id)
identical(example$classification, reloaded$classification)
identical(length(example$horizons), length(reloaded$horizons))
```

## Validation

`validate_soil_profile()` checks the following constraints:

- The object has `soil_profile` class.
- `horizons` is a non-empty list of `soil_horizon` objects.
- Top depths are sorted in ascending order (shallow → deep).
- Each horizon's `bottom > top`.
- No overlapping horizons.

```{r validation}
# Valid profile passes silently
validate_soil_profile(profile)

# Invalid profiles throw informative errors
tryCatch(
  validate_soil_profile(list(not = "a profile")),
  error = function(e) message("Error: ", e$message)
)
```

## Key USDA terminology

Some key terms from the USDA Soil Survey Manual used throughout soilgraph:

- **Horizon**: A layer of soil approximately parallel to the soil surface,
  differing from adjacent layers in physical or chemical properties.
- **Boundary grade (distinctness)**: How sharply one horizon transitions to the
  next — abrupt (< 2 cm), clear (2–5 cm), gradual (5–15 cm), diffuse (> 15 cm).
- **Boundary shape (topography)**: The spatial pattern of the boundary — smooth,
  wavy, irregular, broken, or discontinuous.
- **Coarse fragments**: Rock or mineral particles > 2 mm diameter. Classified by
  size, shape, and degree of weathering (cementation grade).
- **Texture class**: The relative proportion of sand, silt, and clay (e.g. "silt
  loam" = mostly silt with some sand and clay).
